### 2025-07-01 Text/Thesis/Chapters/Chapter5.qmd
Update thesis

>>>>

This thesis presented a statistical framework for detecting heterogeneity in SAR images, offering both theoretical and practical insights into a fundamental question: How can one effectively distinguish heterogeneous clutter from fully developed speckle, assuming SAR intensity is modeled by the Gamma SAR ($\Gamma_{\text{SAR}}$) distribution?

To address this, we proposed three novel hypothesis tests based on entropy measures: Shannon, Rényi, and Tsallis entropies. A key theoretical contribution was the derivation of closed-form analytical expressions for the Rényi and Tsallis entropies under both the $\Gamma_{\text{SAR}}$ and $\mathcal{G}^0_I$ distributions, which were previously unavailable in the literature. Using these expressions, we developed bootstrap-based statistical tests corresponding to each entropy measure.

A comprehensive Monte Carlo simulation study was conducted to evaluate the performance of the proposed tests, assessing both their size (Type&nbsp;I error rate) and power (probability of correctly rejecting the null hypothesis under the alternative). The results demonstrated that the methods maintained reliable control over false alarms while significantly improving detection performance as the sample size and number of looks increased.

These theoretical findings were further validated with experiments on both simulated  and real SAR data acquired by various radar missions (TanDEM-X, PAZ, ICEYE, UAVSAR, Sentinel-1B), covering a range of resolutions, polarizations, and look numbers. Visual analysis using $p$-value heatmaps and binary decision maps revealed notable distinctions among the entropy-based tests. In particular, the Rényi-based test exhibited the highest sensitivity in detecting subtle textural variations, effectively identifying heterogeneity even under high-speckle conditions. The Tsallis-based test provided complementary advantages by reliably identifying homogeneous regions, thereby minimizing false alarms (i.e., incorrectly classifying homogeneous areas as heterogeneous). In contrast, the Shannon-based test displayed intermediate performance and lower robustness at low numbers of looks.

Quantitative evaluations using ground truth regions of interest (ROIs) and standard classification metrics further confirmed the superior overall performance of both Rényi and Tsallis-based tests across all tested scenes.

Compared to conventional classification methods, the proposed entropy-based statistical tests offer practical advantages:  they provide interpretable, analytical significance measures and require no training data or manual labeling, facilitating rapid analysis suitable for environments with limited data availability. Typical computations are efficient, generally taking only a few minutes depending on the image size and number of bootstrap replications. Since the three tests share a common computational structure, their execution times are comparable, making the proposed methods particularly attractive for quick screening and exploratory analysis.

Future research directions identified from this thesis include integrating the improved bootstrap-based Rényi and Tsallis estimators into broader SAR classification frameworks, both supervised and unsupervised, to expand their applicability beyond heterogeneity detection. This integration could enable the development of multi-level texture classification schemes (e.g., low, moderate, high heterogeneity) that go beyond binary discrimination and provide a more detailed understanding of the scene, without requiring labeled training data.

Additionally, extending the proposed methodology to polarimetric SAR (PolSAR) data represents a promising avenue. One approach could involve analyzing the intensity channels independently and fusing the resulting entropy information using multivariate models. Alternatively, a fully polarimetric analysis could be pursued by testing deviations from the scaled complex Wishart distribution. In this context, exploring the asymptotic behavior of the generalized variance (i.e., the determinant of the covariance matrix) under the complex Wishart model may offer valuable theoretical insights for detecting structural changes in polarimetric data.

### 2025-07-01 Text/Thesis/Frontmatter/abstract.tex
Update thesis

%\begin{otherlanguage}{spanish}
%\begin{abstract}
%\phantomsection
%\addcontentsline{toc}{chapter}{Abstract}
This work presents a statistical approach to identify the
underlying roughness characteristics in synthetic aperture radar (SAR)
intensity data. The physical modeling of this kind of data allows the
use of the Gamma distribution in the presence of fully-developed
speckle, i.e., when there are infinitely many independent backscatterers
per resolution cell, and none dominates the return. Such areas are often
called ``homogeneous'' or ``textureless'' regions. The
\(\mathcal{G}_I^0\) distribution is also a widely accepted law for
heterogeneous and extremely heterogeneous regions, i.e., areas where the
fully-developed speckle hypotheses do not hold.
One issue involving the parametric space of $\mathcal{G}_I^0$ is the analytical 
infeasibility of testing homogeneity against heterogeneity using classical tests. 
As solutions to this problem, we propose three test
statistics to distinguish between homogeneous and inhomogeneous regions,
i.e., between Gamma and \(\mathcal{G}_I^0\) distributed data, both with
a known number of looks. The first test statistic uses a bootstrapped
non-parametric estimator of Shannon entropy, providing an
assessment in uncertain distributional assumptions. The second test uses
the classical coefficient of variation (CV). The third test uses an
alternative form of estimating the CV based on the ratio of the mean
absolute deviation from the median to the median. We apply our test
statistic to create maps of \(p\)-values for the homogeneity hypothesis.
Finally, we show that our proposal, the entropy-based test, outperforms
existing methods, such as the classical CV and its alternative variant,
in identifying heterogeneity when applied to both simulated and actual
data.
%\end{abstract}

\par
\noindent \textbf{Keywords:} SAR; heterogeneity; entropy; coefficient of variation; hypothesis tests

>>>>

%\begin{otherlanguage}{spanish}
%\begin{abstract}
%\phantomsection
%\addcontentsline{toc}{chapter}{Abstract}

This work  presents a statistical approach to detect heterogeneity 
in synthetic aperture radar (SAR) intensity data using entropy-based methods.
In homogeneous regions, where fully developed speckle is present, SAR returns 
are well modeled by the Gamma distribution. 
In contrast, heterogeneous areas exhibit more complex scattering behaviors, 
typically modeled by the $\mathcal{G}_I^0$ distribution. Distinguishing between 
these regimes is fundamental for accurate interpretation in remote sensing applications, 
yet classical parametric tests often fail to address this task effectively due to analytical 
and numerical limitations.
To overcome these challenges, we propose three test statistics for detecting heterogeneity 
in SAR imagery based on Shannon, Rényi, and Tsallis entropy measures. These tests rely on 
nonparametric entropy estimators constructed from sample spacings, avoiding explicit assumptions 
about the underlying distribution. To enhance their accuracy, particularly in small samples, 
we incorporate a bootstrap bias-correction procedure that improves the estimators’ stability, 
reduces bias, and lowers mean squared error.
The proposed tests are evaluated using Monte Carlo simulations, assessing size and power 
under varying sample sizes and numbers of looks.
Results demonstrate that the Rényi and Tsallis-based tests outperform the Shannon-based version by 
detecting more subtle texture variations while maintaining greater reliability in identifying 
truly homogeneous regions.
Finally, the methodology is applied to both simulated and real SAR data. We generate $p$-value maps using a 
sliding window analysis, allowing for visual and quantitative assessment of spatial heterogeneity. 
The Rényi-based test consistently identifies fine-scale roughness patterns, 
while the Tsallis-based test shows strength in reliably detecting homogeneous areas. 
Together, these entropy-based tools provide a robust, interpretable, and unsupervised 
framework for heterogeneity detection in SAR data.

\vspace{1em}
\par
\noindent \textbf{Keywords:} SAR; Heterogeneity; Entropy; Nonparametric Estimation; Bootstrap; Statistical Test

### 2025-07-01 Text/Thesis/Frontmatter/resumo.tex
R1 update: revised manuscript and reviewer responses added
Update thesis

\begin{otherlanguage}{portuguese}

\begin{resumo}
\phantomsection
\addcontentsline{toc}{chapter}{Resumo}

Este trabalho apresenta uma abordagem estatística para identificar as características de rugosidade subjacentes em dados de imagens em formation de intensidade de sensores de radar de abertura sintética (SAR -- \textit{Synthetic Aperture Radar}). 
A modelagem física desse tipo de dados permite o uso da distribuição Gamma na presença de speckle totalmente desenvolvido, ou seja, quando há infinitos retroespalhadores independentes por célula de resolução e nenhum domina o retorno.
Essas áreas são frequentemente chamadas de regiões ``homogêneas'' ou ``sem textura''. 
A distribuição $\mathcal{G}_I^0$ também é uma lei amplamente aceita para regiões heterogêneas e extremamente heterogêneas, ou seja, áreas onde as hipóteses de speckle totalmente desenvolvidas não se aplicam.
Uma problemática envolvendo o espaço paramétrico da $\mathcal{G}_I^0$ é a inviabilidade analítica de se testar homogeneidade contra heterogeneidade a partir de testes clássicos. 
Como soluções para esta questão, propomos três testes de hipóteses  para distinguir entre regiões homogêneas e não homogêneas, ou seja, entre dados distribuídos Gamma e $\mathcal{G}_I^0$, ambos com um número conhecido de looks.
A primeira estatística de teste usa um estimador não paramétrico da entropia de Shannon, incorporando metodologias de bootstrap.
O segundo teste usa o coeficiente de variação clássico (CV). 
O terceiro teste usa uma forma alternativa de estimar o CV com base na razão da média do desvio absoluto em relação à mediana.
Mostramos que nossa proposta, o teste baseado em entropia, supera métodos existentes, tais como o CV clássico e sua variante alternativa, na identificação de heterogeneidade quando aplicado a dados simulados e imagens SAR.
Aplicamos nossa estatística de teste para criar mapas de $p$-valores para a hipótese de homogeneidade.

\end{resumo}

\begin{keywords}
SAR; heterogeneidade; entropia; coeficiente de variação; testes de hipóteses
\end{keywords}

\end{otherlanguage}

>>>>

Este trabalho apresenta uma abordagem estatística para a detecção de heterogeneidade 
em dados de intensidade de radar de abertura sintética (SAR) utilizando medidas 
baseadas em entropia. Em regiões homogêneas, onde há speckle totalmente desenvolvido, 
os retornos SAR são bem modelados pela distribuição Gama. Em contraste, áreas heterogêneas 
apresentam comportamentos de espalhamento mais complexos, normalmente descritos pela 
distribuição $\mathcal{G}_I^0$. Distinguir entre esses dois regimes é fundamental 
para uma interpretação precisa em aplicações de sensoriamento remoto. No entanto, 
testes paramétricos clássicos geralmente não são adequados para essa tarefa devido a 
limitações analíticas e numéricas.
Para superar esses desafios, propomos três estatísticas de teste para detectar heterogeneidade 
em imagens SAR com base nas entropias de Shannon, Rényi e Tsallis. Esses testes utilizam 
estimadores não paramétricos de entropia construídos a partir de espaçamentos amostrais, 
evitando suposições explícitas sobre a distribuição subjacente. Para aumentar a precisão dos testes, 
especialmente em amostras pequenas, incorporamos um procedimento de correção de viés via bootstrap, 
que melhora a estabilidade dos estimadores, reduz o viés e o erro quadrático médio.
As estatísticas propostas são avaliadas por meio de simulações de Monte Carlo, 
onde analisamos seu tamanho e poder sob diferentes condições de speckle e textura. 
Os resultados mostram que os testes baseados em Rényi e Tsallis superam a versão baseada 
em Shannon, detectando variações de textura mais sutis e mantendo maior confiabilidade na 
identificação de regiões verdadeiramente homogêneas.
Por fim, a metodologia é aplicada a dados SAR simulados e reais. 
A análise é realizada com janelas deslizantes, gerando mapas de valores de $p$ que permitem a a
valiação visual e quantitativa da heterogeneidade espacial. O teste baseado na entropia de 
Rényi mostra desempenho superior na identificação de padrões de rugosidade em pequena escala, 
enquanto o teste baseado em Tsallis é eficaz na detecção de regiões homogêneas. Em conjunto, 
essas ferramentas baseadas em entropia oferecem uma estrutura robusta, interpretável e não 
supervisionada para a detecção de heterogeneidade em dados SAR.

\vspace{1em}
\par
\noindent \textbf{Palavras-chave:} SAR; Heterogeneidade; Entropia; Estimativa não paramétrica; Bootstrap; Teste estatístico
